\ttitle

### 2013-12-17 Chapters/Chapter1.tex
WIP: Finalizing chapter 3

\section{One-Class Support Vector Machines}
In the setup, as described in the previous section, we employ a model construction phase.
For this model we will use an \acrlong{oc-svm} implementation: the \acrlong{svdd} algorithm by Tax~\cite{tax2001one}.
In earlier research multi-class \glspl{svm} are used for direct classification of inertial sensor data \cite{he2008activity,mountrakis2011support,anguita2012human}.
Others have used \gls{oc-svm} classifiers for novelty, or outlier, detection \cite{scholkopf1999support,camci2010change,li2003improving,ma2003time,tax1999support}.
However, to our knowledge, \gls{oc-svm} methods have not yet been applied to inertial sensor time series data directly.
This algorithm transforms the time series data, which can be of any high dimension, to an higher dimensional feature space.
When the data is assumed to be of the same class, it can be characterized by an enclosing boundary, in the shape of a hypersphere.
Using this \gls{occ} method, we are able to test new data points of being a member of this class.
When a new data point, after transformation to the feature space, lies within the boundary, it is a member of the class, and otherwise it is regarded as an outlier.

@@

\section{Contributions}
As a result from our proposed method and application to recorded real-world data sets, this thesis will provide the following contributions:
\begin{enumerate}
  \item \textbf{Continuous recorded human activity data:} the proposed method is applied to our own recorded human activity data.
  This data set can be grounded by using the video recordings of the subject while performing the activities.
  The data set, consisting of raw inertial sensor recordings, manual change point and activity labeling, and video recordings, are used for testing the method and are made publicly available \cite{vlasveld2013continuous}.
  \item \textbf{Application of \gls{oc-svm} methods to inertial sensor data:} up to our knowledge, this thesis is the first application of \gls{oc-svm} methods, especially the \gls{svdd} method, to inertial sensor signals, such as accelerometer time series data.
  Furthermore, this thesis contributes to the small amount of research that has a temporal segmentation of inertial sensor data as its primary objective.
  \item \textbf{Adaptation of \acrshort{svcpd}:} we have adopted and slightly adapted the \acrshort{svcpd} method by Camci~\cite{camci2010change} in order to find change points in time series data.
  In our simplification we have not encountered noticeably decrease in performance.
\end{enumerate}

>>>>

\section{One-Class Support Vector Machines}
In the setup, as described in the previous section, we employ a model construction phase.
For this model we will use an \acrlong{oc-svm} implementation: the \acrlong{svdd} algorithm by Tax~\cite{tax2001one}.
In earlier research multi-class \glspl{svm} are used for direct classification of inertial sensor data \cite{he2008activity,mountrakis2011support,anguita2012human}.
Others have used \gls{oc-svm} classifiers for novelty, or outlier, detection \cite{scholkopf1999support,camci2010change,li2003improving,ma2003time,tax1999support}.
The research by Yin \etal  \cite{yin2008sensor} creates a \gls{oc-svm} classifier to filter normal activity traces, using training data.
However, to our knowledge, \gls{oc-svm} methods have not yet been applied to inertial sensor time series data directly to find changes in behavior and create a temporal segmentation.
This algorithm transforms the time series data, which can be of any high dimension, to an higher dimensional feature space.
When the data is assumed to be of the same class, it can be characterized by an enclosing boundary, in the shape of a hypersphere.
Using this \gls{occ} method, we are able to test new data points of being a member of this class.
When a new data point, after transformation to the feature space, lies within the boundary, it is a member of the class, and otherwise it is regarded as an outlier.

@@

\section{Contributions}
As a result from our proposed method and application to recorded real-world data sets, this thesis will provide the following contributions:
\begin{enumerate}
  \item \textbf{Continuous recorded human activity data:} the proposed method is applied to our own recorded human activity data.
  This data set can be grounded by using the video recordings of the subject while performing the activities.
  The data set, consisting of raw inertial sensor recordings, manual change point and activity labeling, and video recordings, are used for testing the method and are made publicly available \cite{vlasveld2013continuous}.
  \item \textbf{Application of \gls{oc-svm} methods to inertial sensor data:} to our best knowledge, this thesis is the first application of \gls{oc-svm} methods, especially the \gls{svdd} method, to inertial sensor signals in the context of creating a temporal segmentation.
  Furthermore, this thesis contributes to the small amount of research that has a temporal segmentation of inertial sensor data as its primary objective.
  \item \textbf{Adaptation of \acrshort{svcpd}:} we have adopted and slightly adapted the \acrshort{svcpd} method by Camci~\cite{camci2010change} in order to find change points in time series data.
  In our simplification we have not encountered noticeably decrease in performance.
\end{enumerate}

### 2013-12-17 Chapters/Chapter2/svm.tex
Cleanup of sections
WIP: Finalizing chapter 3

% !TEX root = ../../main.tex
\section{Support Vector Machines in Change Detection}\label{sec:svm_in_change_detection}

@@

The method by Yin and Yang \cite{yin2008sensor} uses the \gls{svdd} method in the first phase of a two-phase algorithm to filter commonly available normal activities.
The \gls{svcpd} method of Camci \cite{camci2010change} uses \gls{svdd} applies it to time-series data.
By using an \gls{oc-svm} the method is able to detect changes in mean variance.
Especially the detection of variance \emph{decrease} is an improvement over other methods that rely in the \gls{kl} divergence, since these are unable to detect decreases in variance \cite{takeuchi2006unifying}.
This main research subject of this thesis is to apply the method of Camci \cite{camci2010change} to sensor data (such as accelerometer time series) obtained by on-body smartphones.

>>>>

% !TEX root = ../../main.tex
\section{Support Vector Machines in Change Detection}\label{sec:literature_review_svm}

@@

The method by Yin \etal \cite{yin2008sensor} uses the \gls{svdd} method in the first phase of a two-phase algorithm to filter commonly available normal activities.
The \gls{svcpd} method of Camci \cite{camci2010change} uses \gls{svdd} applies it to time-series data.
By using an \gls{oc-svm} the method is able to detect changes in mean variance.
Especially the detection of variance \emph{decrease} is an improvement over other methods that rely in the \gls{kl} divergence, since these are unable to detect decreases in variance \cite{takeuchi2006unifying}.
This main research subject of this thesis is to apply the method of Camci \cite{camci2010change} to sensor data (such as accelerometer time series) obtained by on-body smartphones.

### 2013-12-17 Chapters/Chapter3.tex
Cleanup of sections
WIP: Finalizing chapter 3

%----------------------------------------------------------------------------------------
\section{Outline}
\emph{Not intented for the reader.}
\begin{itemize}
  \item Content follows blogpost, but more formal
  \item Explain the connection between change detection and \{outlier detection, density estimation, \etc\}
  \item Explain change detection in relation with ``M-Estimators''.
  \item Explain why and how SVM can be used for outlier detection, and thus for change detection in time series
  \item Explain options when using SVM, such as RBF kernel
  \item Relate characteristics of (accelerometer) data to the used (RBF?) kernel
  \item Define some quality metrics --> this should be in a different chapter (About experiments)
  \item Note: try to avoid the perspective of \emph{density estimation}. Instead: \emph{data description} or \emph{boundary description}.
\end{itemize}

This chapter discusses the concepts and algorithms that will be used as a basis for the proposed method, as discussed in \Cref{Chapter4}.
The first sections formulates the problem of change detection and relates it to outlier and novelty detection.
It transforms the traditional problem to change detection for times series data.
In \Cref{sec:one_class_classification} the problem of \acrlong{occ} is discussed and shows a number of implementations.
Two \acrlong{svm}-based methods, \gls{nu-svm} and \gls{svdd}, are further disussed in \Cref{sec:one_class_svm}.
For the application to the problem of \acrlong{har}, the characteristics of the collected sensor data is discussed in~\ref{sec:sensor_data_characteristics}.
Finally a theoretic set of quality metrics is discussed in \Cref{sec:quality_metrics}.

\input{./Chapters/Chapter3/change_detection_time_series}
\input{./Chapters/Chapter3/one_class_classification}
\input{./Chapters/Chapter3/one_class_svm}
\input{./Chapters/Chapter3/sensor_data_characteristics}
\input{./Chapters/Chapter3/quality_metrics}

>>>>

This chapter discusses the concepts and algorithms that will be used as a basis for the proposed method, as introduced in \Cref{Chapter4}.
The first section formulates the problem of change detection and relates it to outlier and novelty detection.
It transforms the traditional problem of outlier detection into change detection for times series data.
In \Cref{sec:one_class_classification} the problem of \acrlong{occ} is explained and discusses a number of implementations.
In the final sections of this chapter, \Cref{sec:one_class_svm}, two \gls{svm}-based \gls{occ} methods, \gls{nu-svm} and \gls{svdd}, and the influence of model parameters are further discussed in detail.

\input{./Chapters/Chapter3/change_detection_time_series}
\input{./Chapters/Chapter3/one_class_classification}
\input{./Chapters/Chapter3/one_class_svm}

### 2013-12-17 Chapters/Chapter3/change_detection_time_series.tex
WIP: Finalizing chapter 3

% !TEX root = ../../main.tex
\section{Change Detection in Time Series}\label{sec:change_detection_time_series}
\TODO{Write section introduction} \\
\emph{The section should relate the problem of outlier detection to change detection.
It should transform the traditional problem formulation to that of time series data.
Build to one-class classification, which can be used for outlier detection}

\subsection{Relation to outlier detection}
The problems of outlier and novelty detection, segmentation, and change detection are closely related.
The terminology depends on the field of application, but there are subtle differences.
The problem of outlier detection is concerned with finding data objects in a data set which have small resemblance with the majority of the data objects.
These objects can be regarded as erroneous measurements.
In the case of novelty detection these objects are considered to be member of a new class of objects.
The unifying framework of Takeuchi and Yamanishi \cite{takeuchi2006unifying}, \gls{changeFinder}, creates a two stage process expressing change detection in terms of outlier detection.
The first stage determines the outliers in a time series by giving a score based on the deviation from a learned model, and thereby creates a new time series.
The second stage runs on that new created time series and calculates a average over a window of the outlier scores.
The problem of change detection is then reduced to outlier detection over that average-scored time series.
The implementation by Camci \cite{camci2010change}, \gls{svcpd}, implements outlier detection with the \gls{svdd} algorithm to detect changes in mean and variance.

\subsection{Problem formulation}\label{subsec:change_detection_problem_formulation}
The problem of change point detection can be formulated using different type of models, as discussed in~\ref{subsec:segmentation}.
The methods by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and Camci \cite{camci2010change} use the following formulation for change detection, which we will also use for our problem formulation.
The algorithm searches for sudden changes in the time series data.
In other words, slowly changing properties in the data are not considered to be changes.
Considered a time series $x_1 x_1 \dots$, which is drawn from a stochastic process $p$.
Each $x_t$ (t = 1, 2, \dots) is a $d$-dimensional real valued vector and $p$ a probability density function of the sequence $x_1 x_2 \dots$.
Assume $p$ can be decomposed in two different \gls{iid} stationary stochastic processes $p^1$ and $p^2$ and are one-dimensional Gaussian density functions.
For a time point $a$ data points for which $t < a$ are drawn from $p^1 = N(\mu_1, \sigma_1^2)$ and for $t \geq a$ from $p^2 = N(\mu_2, \sigma_a^2)$.
If $p^1$ and $p^2$ are different, then the time point $t = a$ is a \emph{change point}.
In \cite{takeuchi2006unifying} the similarity between the stochastic processes are expressed by the \gls{kl} divergence $D(p^2||p^1)$.
It is observed that this measure is not able to detect a change by decrease in variance \cite{takeuchi2006unifying,camci2010change}.
This problem is the motivation for Camci \cite{camci2010change} to create the \gls{svcpd} algorithm.

The definition of change point being sudden changes in the time series data is in line with the search of changes in activities.
Since we are only interested in different activities (which are represented by sudden changes), slight changes within an activity are not of interest.
\Cref{sec:sensor_data_characteristics} discusses the representation of activities and changes between them in the data in more detail.

\TODO{add more formal definition or analysis of change detection}

-- Literature --

``Online segmentation of time series based on polynomial least-squares approximations'', \cite{fuchs2010online}. 25, 2010 \\

>>>>

% !TEX root = ../../main.tex
\section{Problem Formulation Change Detection}\label{sec:change_detection_time_series}
% \TODO{Write section introduction} \\
% \emph{The section should relate the problem of outlier detection to change detection.
% It should transform the traditional problem formulation to that of time series data.
% Build to one-class classification, which can be used for outlier detection}

% \subsection{Relation to outlier detection}
The problems of outlier and novelty detection, segmentation, and change detection in time series data are closely related.
The terminology depends on the field of application, but there are subtle differences.
The problem of outlier detection is concerned with finding data objects in a data set which have small resemblance with the majority of the data objects.
These objects can be regarded as erroneous measurements.
In the case of novelty detection these objects are considered to be member of a new class of objects.
The unifying framework of Takeuchi and Yamanishi \cite{takeuchi2006unifying}, \gls{changeFinder}, creates a two stage process expressing change detection in terms of outlier detection.
The first stage determines the outliers in a time series by giving a score based on the deviation from a learned model, and thereby creates a new time series.
The second stage runs on that new created time series and calculates a average over a window of the outlier scores.
The problem of change detection is then reduced to outlier detection over that average-scored time series.
The implementation by Camci \cite{camci2010change}, \gls{svcpd}, implements outlier detection with the \gls{svdd} algorithm to detect changes in mean and variance.

% \subsection{Problem formulation}\label{subsec:change_detection_problem_formulation}
The problem of change point detection can be formulated using different type of models, as discussed in~\ref{subsec:segmentation}.
The methods by Takeuchi and Yamanishi \cite{takeuchi2006unifying} and Camci \cite{camci2010change} use the following formulation for change detection, which we will also use for our problem formulation.
The algorithm searches for sudden changes in the time series data.
In other words, slowly changing properties in the data are not considered to be changes.
Considered a time series $x_1 x_1 \dots$, which is drawn from a stochastic process $p$.
Each $x_t$ (t = 1, 2, \dots) is a $d$-dimensional real valued vector and $p$ a probability density function of the sequence $x_1 x_2 \dots$.
Assume $p$ can be decomposed in two different \gls{iid} stationary stochastic processes $p^1$ and $p^2$ and are one-dimensional Gaussian density functions.
For a time point $a$ data points for which $t < a$ are drawn from $p^1 = N(\mu_1, \sigma_1^2)$ and for $t \geq a$ from $p^2 = N(\mu_2, \sigma_a^2)$.
If $p^1$ and $p^2$ are different, then the time point $t = a$ is a \emph{change point}.
In \cite{takeuchi2006unifying} the similarity between the stochastic processes are expressed by the \gls{kl} divergence $D(p^2||p^1)$.
It is observed that this measure is not able to detect a change by decrease in variance \cite{takeuchi2006unifying,camci2010change}.
This problem is the motivation for Camci \cite{camci2010change} to create the \gls{svcpd} algorithm.

The definition of change point being sudden changes in the time series data is in line with the search of changes in activities.
Since we are only interested in different activities (which are represented by sudden changes), slight changes within an activity are not of interest.

% -- Literature --

% ``Online segmentation of time series based on polynomial least-squares approximations'', \cite{fuchs2010online}. 25, 2010 \\

### 2013-12-17 Chapters/Chapter3/one_class_classification.tex
WIP: Finalizing chapter 3

% !TEX root = ../../main.tex
\section{One-Class Classification}\label{sec:one_class_classification}
\TODO{Add ref to \cite{hodge2004survey}, which states 3 types of outlier detection. It puts \gls{occ} as of Type 3: \textbf{semi-supervised recognition or detection}}

As discussed in the previous section, change detection in time series can be implemented using outlier, or novelty, detection.
To regard a data point as an outlier, there must be a model of the normal time series data and a (dis)similiarity measure defined over the model and data objects.
When a data point differs enough from the created model, it can be labeled as an outlier.
The class of \gls{occ} algorithms is especially designed for that purpose.
The algorithms build up a model of the data, assuming only normal data objects are available (or a very limited amount of example outlier data objects).
This differs from classical classification algorithms, which commonly rely of both positive and negative examples.

@@

The \gls{occ} algorithms have been applied to a wide range of applications.
The first is, obviously, outlier detection of objects which do not resemble the bulk of the training data.
It can be a goal by itself and can be used as a filtering mechanism for other data processing methods.
Often methods that rely on data-characteristics are sensitive for remote regions in the data set.
Using \gls{occ} these remote regions can be removed from the data set.
A second application is for the problem as described above, in which only data from a single target class is available.
When this problems originates from \eg a monitoring process, the \gls{occ} is able to recognize abnormal system states, without the need to create (or simulate) faulty states beforehand.
The final possible application given by Tax \cite{tax2001one} is the comparison of two data sets.
By constructing a \gls{occ}-classifier using a training data set, one can compare that set to a new data set.
This will result in a similarity-measure expressed in the properties of outliers.
This measure is related to other similarity measures, such as \gls{pdf} estimation and the \gls{kl} divergence as discussed in \Cref{subsec:change_detection_problem_formulation}.

@@

In this Section we have discussed the problem of \gls{occ} and different kind of implementations.
An often used implementation is the \gls{svm}-based method \cite{noumir2012simple}, since is shows good performance in comparitive researches.
\TODO{Add refs for good results of OC-SVM}
In the following section (\ref{sec:one_class_svm}) two implementations of \gls{oc-svm} will be discussed, the \gls{svdd} method of Tax and Duin \cite{tax1999support} and the \gls{nu-svm}-algoritm by Sch\"olkopf.

>>>>

% !TEX root = ../../main.tex
\section{One-Class Classification}\label{sec:one_class_classification}
As discussed in the previous section, change detection in time series can be implemented using outlier, or novelty, detection.
To regard a data point as an outlier, there must be a model of the normal time series data and a (dis)similiarity measure defined over the model and data objects.
When a data point differs enough from the created model, it can be labeled as an outlier.
The class of \gls{occ} algorithms is especially designed for that purpose.
The algorithms build up a model of the data, assuming only normal data objects are available (or a very limited amount of example outlier data objects).
This is also known as novelty detection or semi-supervised detection and is of Type 3 in the system by Hodge and Austin~\cite{hodge2004survey}.
This differs from classical classification algorithms, which commonly rely of both positive and negative examples.

@@

The \gls{occ} algorithms have been applied to a wide range of applications.
The first is, obviously, outlier detection of objects which do not resemble the bulk of the training data.
It can be a goal by itself and can be used as a filtering mechanism for other data processing methods.
Often methods that rely on data-characteristics are sensitive for remote regions in the data set.
Using \gls{occ} these remote regions can be removed from the data set.
A second application is for the problem as described above, in which only data from a single target class is available.
When this problems originates from \eg a monitoring process, the \gls{occ} is able to recognize abnormal system states, without the need to create (or simulate) faulty states beforehand.
The final possible application given by Tax \cite{tax2001one} is the comparison of two data sets.
By constructing a \gls{occ}-classifier using a training data set, one can compare that set to a new data set.
This will result in a similarity-measure expressed in the properties of outliers.
This measure is related to other similarity measures, such as \gls{pdf} estimation and the \gls{kl} divergence as discussed in \Cref{subsec:change_detection_time_series}.

@@

In this Section we have discussed the problem of \gls{occ} and different kind of implementations.
An often used implementation is the \gls{svm}-based method \cite{noumir2012simple}, since is shows good performance in comparitive researches \cite{khan2010survey,smola1998connection}.
In the following section (\ref{sec:one_class_svm}) two implementations of \gls{oc-svm} will be discussed, the \gls{svdd} method of Tax and Duin \cite{tax1999support} and the \gls{nu-svm}-algoritm by Sch\"olkopf.
